Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-25 paper/paper.tex
Begin Technique and Results subsection under Rating-based Method.

As already mentioned, the estimator $PearSim$ is unable to say much about the
similarity of two items when only provided ratings from a single common user. In
this case, $PearSim$ yields $\pm1$ (or the indeterminate value $0/0$ in the case
that both ratings are zero) depending on whether the ratings have the same sign
or not. The value of $\hat{y}$ provided by the above model rectifies this by
generating similarity estimates in the range $[-1, +1]$ given a single pair of
ratings. We therefore consider each user who rates a pair of items as implicitly
predicting the similarity of the items. A key discovery of the present work is
that the accuracy of users' predictions in a training set is correlated with the
accuracy in a test set. We exploit this finding by ascribing a quantity we call
{\em user predictivity} to each user defined as
\begin{align}
p_u = e^{-\epsilon_u/\lambda},
\end{align}
where $\lambda$ is a parameter to be learned and
\begin{align}
\epsilon_u = \frac{1}{\left|I_u\right|}
\sum_{A,B \in I_u}{\left|ModelSimRand(A, B) - GoldSimRand(A, B)\right|}
\end{align}
with $I_u$ the set of item pairs rated by user $u$. We are once again using
$GoldSim$ as a proxy for $TrueSim$ (See the {\bf The Experiment} section) such
that $\epsilon_u$ is the average absolute error of user $u$'s similarity
predictions.

@@

@@

>>>>

As already mentioned, the estimator $PearSim$ is unable to say much about the
similarity of two items when only provided ratings from a single common user. In
this case, $PearSim$ yields $\pm1$ (or the indeterminate value $0/0$ in the case
that both ratings are zero) depending on whether the ratings have the same sign
or not. The value of $\hat{y}$ provided by the above model rectifies this by
generating similarity estimates in the range $[-1, +1]$ given a single pair of
ratings. We therefore consider each user who rates a pair of items as implicitly
predicting the similarity of the items. A key discovery of the present work is
that the accuracy of users' predictions in a training set is correlated with the
accuracy in a test set. We exploit this finding by ascribing a quantity we call
{\em user predictivity} to each user defined as
\begin{align}
p_u = e^{-\epsilon_u/\lambda},
\end{align}
where $\lambda$ is a parameter to be learned and
\begin{align}
\epsilon_u = \frac{1}{\left|I_u\right|}
\sum_{A,B \in I_u}{\left|ModelSimRand(A, B) - GoldSimRand(A, B)\right|}
\end{align}
with $I_u$ the set of item pairs rated by user $u$. We are once again using
$GoldSim$ as a proxy for $TrueSim$ (See the {\bf Experiment} section) such that
$\epsilon_u$ is the average absolute error of user $u$'s similarity predictions.

@@

The parameter values $\mu_s$ and $\sigma_s$ are obtained by fitting the
distribution of $GoldSimRand$ values in the training set (see Fig.
\ref{fig:Histograms}). For example, the $N \geq 80$ dataset we have $\mu_s =
0.27$ and $\sigma_s = 0.25$. The choice of $\mu_r$ and $\sigma_r$ is less clear.
Some insight can be gained, however, by examining the prior distribution of
ratings for our datasets, which are shown in Fig. \ref{fig:RatingDist}.

@@

The upper panel of Fig. \ref{fig:RatingDist} shows data for the $N \geq 40$
training set while the lower panel shows the same for the $N \geq 80$ training
set. In these density plots, for each rating pair, we randomly choose which is
Rating1 and which is Rating2. The two density plots present similar
distributions. These data, together with the similarity score histograms of Fig.
\ref{fig:Histograms}, suggest that the data for our two data sets are similar
and that large biases are not introduced by varying the number of common users,
$N$. At the same time, both rating distributions reveal significant correlations
between Rating1 and Rating2 as evidenced by the elongation in the $45^\circ$
direction. This is consistent with the data of Fig. \ref{fig:Histograms}, which
shows the mean $PearSim > 0$ for both sets. Further, these data indicate a
bias toward positive values among rating pairs compared to those of ratings in
general, which have an average value of zero by construction\footnote{See
discussion of bias subtraction in {\bf Data} section.}. This latter feature
indicates that items with large numbers of common users tend to be rated higher
by their common users than by their other users since we have already subtracted
the item's average rating thereby avoiding any simple popularity effects.

We observe that while the data of Fig. \ref{fig:RatingDist} clearly do not
conform precisely to a multivariate Gaussian distribution, this approximation is
not an unreasonable one.
